---
title: "Did the 1977 Daphne Major drought drive natural selection in *Geospiza fortis*?"
format: html
editor: visual
---

## Introduction

The aim of this analysis is to test whether the beak depth of the medium ground finch (*Geospiza fortis*) on the Galápagos island of Daphne Major changed in response to a severe drought, which altered food availability and potentially imposed a selection pressure. Specifically, we will analyse beak depths (depth of the beak at its base in mm) of 89 individuals, caught before and after the 1977 drought and determine if there is a significant difference between measurements.

## 1. Methods – Statistical Analysis

I perform a linear model analysis (one way ANOVA) as the data is a continuous response variable and a categorical explanatory variable - a factor with two levels (1976, 1978). Fully balanced (equal number of observations for both levels as same individuals) so good for ANOVA as ensure homogeneity of variances? A short paragraph explaining the data, statistical model and analysis

#### Loading Packages

```{r}
#| label: Making a Box Plot
library(here)
library(Sleuth3)
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(car)
library(sjPlot)
library(patchwork)

#Getting the data, converting year into a factor
beak_data <- case0201
str(case0201)
summary(case0201)
beak_data$Year <- as.factor(beak_data$Year)
```
#### Graphing the raw data
```{r}
#Making a colour palette that is more colour blind friendly
custom_colour_palette <- c("1976" = "#E82069", "1978" = "#EFB300")

#Plotting a boxplot with jitter points, adding a legend and adjusting font sizes
ggplot(data = beak_data, aes(x = Year, y = Depth)) + 
  geom_boxplot(aes(color = Year), show.legend = FALSE) +
  geom_jitter(aes(color = Year), 
              show.legend = FALSE, 
              position = position_jitter(width = 0.2, seed = 0)) +
  labs(title = "Figure 1: Comparison of Beak Depths in 1976 vs 1978", 
       x = "Year", 
       y = "Beak Depth (mm)",
       caption = "Base of beak depth data collected from 89 individuals before (1976) and after (1978) the 1977 drought.") + 
  scale_color_manual(values = custom_colour_palette) +  
  theme(
    plot.title = element_text(size = 16, face = "bold"),  
    axis.title.x = element_text(size = 14),  
    axis.title.y = element_text(size = 14),  
    plot.caption = element_text(size = 11),
    axis.text.x = element_text(size = 11),  
    axis.text.y = element_text(size = 11))
```

-   A figure displaying the key result of the linear model analysis.
-   A multi-panel figure allowing graphical assessment of the normality and heteroscedasticity of the residuals

Note also need Table(s): Include a small table of some appropriate output generated by your analysis

#### Checking assumptions and graphing them

```{r}
#|label: Assessing the normality and heteroscedasticity of the residuals

#Make linear model 
beak_depth_linear_model <- lm(Depth ~ as.factor(Year), data = beak_data)

#Extract fitted and residual values, make a dataframe for them 
beak_depth_linear_model_fitted <- fitted(beak_depth_linear_model)
beak_depth_linear_model_residuals <- residuals(beak_depth_linear_model)

residual_fitted_data <- data.frame(
  Fitted = beak_depth_linear_model_fitted,
  Residuals = beak_depth_linear_model_residuals
)

#Plotting residuals against a normal distribution to test for normality (QQ Plot) - shows a lot of deviation so likely non normal 
qq_plot <- ggplot(residual_fitted_data, aes(sample = Residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "Normal Q-Q Plot",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal()

##Shapiro test - statistically tests for normality of residuals - significant result so NOT! normally distributed.
shapiro.test(residual_fitted_data$Residuals)

#Making a plot to assess heteroscedacity of the residuals - residual plot. Does suggest there may be heteroscedacity.
residual_plot <- ggplot(residual_fitted_data, aes(x = Fitted, y = Residuals)) +
  geom_point() +  # Plot the points
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +  # Add horizontal line at 0
  labs(
    title = "Residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_minimal()  

multi_panel_figure <- qq_plot + residual_plot + 
  plot_annotation(title = "Assessing Normality and Heteroscedasticity of Residuals")
multi_panel_figure

#Statistical test of heteroscedacity of residuals (equal variance) - Levene's Test. Non significant result so homoscedacity of residuals 
leveneTest(beak_depth_linear_model_residuals ~ as.factor(Year), data = beak_data)

```

```{r}
#|label: Analysing the linear model

# Display summary of linear model - and turn into a nice table
summary(beak_depth_linear_model)
tab_model(beak_depth_linear_model,
  show.p = TRUE,
  show.ci = TRUE ,
  dv.labels = "Beak Depth",
  pred.labels = c("Intercept (1976)", "1978"))


```


## 2. Results

Mean beak depth in 1976 was 9.47mm and 10.14mm in 1978, an increase of 0.67mm, which was statistically signficant.

```{r}
#|label: reporting output of regression


```

A short paragraph reporting the results of the analysis

## 3. Conclusion

To conclude, the results show that there is a signficant difference between beak depths in 1976 and 1978. Maybe some other explanation - small sample size?

A concise conclusion with respect to the main aims and questions.



Where I left off - need to figure out how to make a graph showing the main result of linear regression 
Need to write up text sections
Need to make table look better
Need to figure out how to make the quarto doc look better in general as a lot of noise
